---
title: "R_Cyclistic_Business_Full_Year_Analysis"
author: "Michael"
date: "2021/9/22"
output: html_document
---
# 1. Project Backgroud and working environment setting
## Project background introduction
Cyclistic bike-share analysis is a frictional case study organized by Google Data Analytics capstone project team. In the project hypothesis, the marketing director of Cyclistic company, that is Lily Moreno, what to design a marketing strategy for converting casual riders to annual members to reinforce company’s future. The marketing strategy should answer the following questions.

1. How do annual members and casual riders use Cyclistic bikes differently? 
2. Why would casual riders buy Cyclistic annual memberships? 
3. How can Cyclistic use digital media to influence casual riders to become members?

This section analysis is trying to answer the first question, finding behavior insights or patterns between two type customers. 

## Data resources and ROCCC inspect
1. Cyclistic is a fictional company for the purposes of this case study. 
2. The data has been made available by Motivate International Inc. under [this license](https://www.divvybikes.com/data-license-agreement)
3. The data was downloaded from amazonaws website [link](https://divvy-tripdata.s3.amazonaws.com/index.html).
4. As case study statement required, downloaded previous 12 months data, that’s from Sep 2020 to Aug 2021.

## Install required packages and other environments setting.
```{r environment seeting, echo=FALSE, warning=FALSE}
Sys.setlocale("LC_TIME", "English")
library(tidyverse)
library(ggplot2)
library(lubridate)
```


# 2. Inspecting monthly Data
## Ispecting data of August 2021.
```{r upload and check data, echo=TRUE, warning=FALSE}
setwd("D:/resume_portfolio/cyclistic_google_case1/divvy_tripdata")  
m8_2021 <-read_csv('202108-divvy-tripdata.csv')  #Rows: 804352 Columns: 13 
colnames(m8_2021)
str(m8_2021)
```
## Uploading other 11 months' data.
```{r upload 11 months data, echo=FALSE, warning=FALSE}
setwd("D:/resume_portfolio/cyclistic_google_case1/divvy_tripdata")
m7_2021 <-read_csv('202107-divvy-tripdata.csv')
m6_2021 <-read_csv('202106-divvy-tripdata.csv')
m5_2021 <-read_csv('202105-divvy-tripdata.csv')
m4_2021 <-read_csv('202104-divvy-tripdata.csv')
m3_2021 <-read_csv('202103-divvy-tripdata.csv')
m2_2021 <-read_csv('202102-divvy-tripdata.csv')
m1_2021 <-read_csv('202101-divvy-tripdata.csv')
m12_2020 <-read_csv('202012-divvy-tripdata.csv')
m11_2020 <-read_csv('202011-divvy-tripdata.csv')
m10_2020 <-read_csv('202010-divvy-tripdata.csv')
m9_2020 <-read_csv('202009-divvy-tripdata.csv')

```
## Problem solving: 
1. start_station_id and end_station_id have different datatype from previous months. 
November, October and September of 2020
```{r Change data types, echo=FALSE}
m11_2020 <- transform(m11_2020, start_station_id=as.character(start_station_id), 
                      end_station_id=as.character(end_station_id))
m10_2020 <- transform(m10_2020, start_station_id=as.character(start_station_id), end_station_id=as.character(end_station_id))
m9_2020 <- transform(m9_2020, start_station_id=as.character(start_station_id), end_station_id=as.character(end_station_id))

```

# 3. Stack and inspect, process Data

### Stack individual month's data frames into whole year data frame. And delete unnecessary features. 
```{r combine data, echo=FALSE, message=FALSE, warning=FALSE}
trips_y <- bind_rows(m8_2021, m7_2021, m6_2021, m5_2021, m4_2021, m3_2021, m2_2021, m1_2021, 
                     m12_2020, m11_2020, m10_2020, m9_2020)
trips_y <- trips_y %>%
  select(-c(start_lat,start_lng,end_lat,end_lng))
head(trips_y)
```
### Inspect new table, Summary for numeric and datetime columns
```{r inspect new table, echo=FALSE, message=FALSE, warning=FALSE}
colnames(trips_y)  #list of column names
nrow(trips_y) # get rows of the table  4.91M
dim(trips_y)
head(trips_y)
str(trips_y)
summary(trips_y) 
```

### Character columns: More detailed information
```{r check character columns, echo=FALSE, message=FALSE, warning=FALSE}
length(unique(trips_y$ride_id))  #4912863
unique(trips_y$rideable_type)  #"electric_bike" "classic_bike"  "docked_bike"  
table(trips_y$member_casual)  # casual 2225089  member 2687983. Almost equal distribution.
length(unique(trips_y$start_station_name))  # 758
length(unique(trips_y$start_station_id))  #1294
length(unique(trips_y$end_station_name))  #757
length(unique(trips_y$end_station_id))  #1294
```

### check missing values for each columns
```{r missing value, echo=FALSE, message=FALSE, warning=FALSE}
for (col in names(trips_y)) {
  missing <- sum(is.na(trips_y[,col]))
  if (missing > 0) {
    print(c(col, missing))
          }
}
```

results
1. "start_station_name" "450045" NA         
2. "start_station_id" "450571"  NA         
3. "end_station_name" "491380"  NA        
4. "end_station_id" "491764" NA

### Increase columns for better presenting data. trip_duration, year, month, day, weekday
```{r increase columns, echo=FALSE, message=FALSE, warning=FALSE}
trips_y$year <- format(trips_y$started_at,"%Y")
trips_y$month <- format(trips_y$started_at,"%m")
trips_y$day <- format(trips_y$started_at,"%d")
trips_y$day_of_week <- format(trips_y$started_at,"%A")

```

increase duration and change datatype
```{r}
trips_y$trip_duration <- difftime(trips_y$ended_at, trips_y$started_at)

trips_y$trip_duration <- as.numeric(as.character(trips_y$trip_duration))
```

### Delete records which hold invalid value in trip_duration column.
```{r delete invalid value}
trips_y2 <- trips_y[trips_y$trip_duration >= 0,]

```

### Order day_of_week.
```{r Order day_of_week, echo=FALSE, message=FALSE, warning=FALSE}
trips_y2$day_of_week <- ordered(trips_y2$day_of_week, 
                                levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

# 4. Conduct data analysis and create visuals.
### 01 Business increase during the last 12 months.
conclusion: Casual users increased faster than membership users during the previous 12 months.

```{r monthly increase, echo=TRUE, message=FALSE, warning=FALSE}
trips_y2$year_month <- format(trips_y2$started_at,"%Y-%m")

trips_y2 %>%
  group_by(member_casual, year_month) %>%
  summarise(number_of_rides = n(), average_duration = mean(trip_duration)) %>%
  arrange(member_casual,year_month) %>%
  ggplot(aes(x=year_month, y=number_of_rides, fill=member_casual)) +
  geom_col(position='dodge')
```

### 02 The number of rides vs weekday and rider types
conclusion: No obvious difference found between weekdays and weekends.

```{r weekday analysis between 2 types, message=FALSE, warning=FALSE}
trips_y2 %>%
  mutate(weekday = wday(trips_y2$started_at, label = TRUE)) %>%    #increase a column, get weekday
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(), average_duration = mean(trip_duration)) %>%  #calculates number of rides
  arrange(member_casual,weekday) %>%
  ggplot(aes(x=weekday, y=number_of_rides, fill=member_casual)) +
  geom_col(position='dodge')
```

### 03 The number of rides vs weekday and rider types
conclusion: Casual users' average duration is more than 2 times of membership ones.

```{r duration vs weekday and customer type, echo=TRUE, message=FALSE, warning=FALSE}
trips_y2 %>%
  mutate(weekday = wday(trips_y2$started_at, label = TRUE)) %>%    #increase a column, get weekday
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(), average_duration = mean(trip_duration)) %>%  #calculates number of rides
  arrange(member_casual,weekday) %>%
  ggplot(aes(x=weekday, y=average_duration, fill=member_casual)) +
  geom_col(position='dodge')
```

### 04 Who made long time (greater than 1 hour) rides?
conclusion: Casual users dominate long time (> 1 hour) rides.

```{r}
trips_y2 %>%
  filter(trip_duration > 3600) %>%
  group_by(member_casual, day_of_week) %>%
  summarise(number_of_rides = n(), average_duration = mean(trip_duration)) %>%
  arrange(member_casual,day_of_week) %>%
  ggplot(aes(x=day_of_week, y=number_of_rides, fill=member_casual)) +
  geom_col(position='dodge')
```

# 5. Export summary for future usage
```{r export, echo=FALSE, message=FALSE, warning=FALSE}
counts <- aggregate(trips_y2$trip_duration~trips_y2$member_casual+trips_y2$day_of_week, FUN = mean)
setwd('D:/resume_portfolio/cyclistic_google_case1')
write.csv(counts,file='trip2.csv')
```

